Making the Best of a Bad Situation

### 2022-03-18 paper.tex
making the best

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Finding Good Generators with Multi-Armed Bandits}

@@

\section{Motivation}
Deductive Program Verifiers like Dafny and Frama-C have proven themselves both useful and powerful tools in the fight against un-verified software.
However, their adoption is hindered by the fact that graduate-level expertise in verification is required in order to verify any non-trivial program. One approach to lowering the barrier to entry for deductive verifiers is to incorporate lighter-weight formal methods into the mix. While SMT based verification is very powerful, its nuances lead to much of the difficulty in using verifiers based on it, and so selectively substituting this verification technique with a less precise but easier-to-use formal method would lower the difficulty while retaining the same workflow and some of the theoretical guarantees.

In this report, we investigate the use of \textit{Property-Based Testing} in place of SMT solving as a core for program verification tools. Property-Based Testing (PBT) \cite{qc} is a lightweight formal method wherein the specifications for a program are checked by randomly generating thousands of inputs, and ensuring that the program meets its specification by executing it at those inputs. While PBT doesn't give full correctness guarantees like SMT solving does, it brings a level of confidence in a manner similar to Model Checking. Moreover, PBT is fully modular, and so it can be used selectively: randomly-tested code and SMT-verified code can coexist in the same file, enabling users to take a \textit{gradual verification} approach by gradually replacing tested code with verified code.

Further, PBT has significant usability benefits over SMT-based verification. A common complaint while using SMT-based verifiers is that it is often difficult to interpret the SMT output. When an SMT solver fails to verify a specification, it can be for one of two reasons: either (a) the specification is incorrect, or (b) the solver was simply unable to find a proof. These two results are often indistinguishable, as the solver may not return a counter-model indicating that the specification is falsifiable --- in this case, the only course of action is to attempt to provide more hints to the solver in the hopes that it will eventually prove the property. With PBT, the only failure mode is through counter-examples, and so no interpretation is required. Moreover, the counter-examples in a PBT approach are more easily interpretable. With SMT, counter-examples are encoded in an SMT data format and are not easily lifted back to program values. With testing, the inputs were concrete to begin with, and so counter-examples can be presented as-is. Finally, PBT can actually aid in the development of SMT-verified programs by allowing the user to test specifications to gain confidence that they are true before spending costly time attempting to develop tricky loop invariants.

In this report, we develop a simple deductive program verifier based on Dafny which is backed by PBT instead of SMT. In \autoref{sec:pbt}, we show how PBT can be used in place of an SMT solver to do deductive verification. This process is conceptually simple, but we will see that it hides a major challenge: generating thousands of independent random input sets which satisfy a function's precondition is NP-hard. In fact, tackling a small variant of problem is the bulk of the content of this report, and its main technical contribution. We give a sketch of this contribution in this section. In brief, our technique first infers a set of ``candidate" example-generators from the precondition in question, and then uses a reinforcement-learning-based online learning algorithm to find the best generator among the set.

@@

%In this report, I show how present a research direction (and preliminary results) which aims to lower the barrier to entry for using verifiers like Dafny.

%At present, there is a great deal of active research into ways to lower the barrier to entry for using verifiers like Dafny. In this report, I present

@@

In terms of ($\ast$), this means that a user of a verifier backed by PBT would have to provide programs which generate random heaps $\sigma$ to satisfy the preconditions of every function they plan to test. This is labor-intensive, error prone, and diminishes the benefits of PBT over SMT-based verification. To make the former practical, we need to \textit{automatically} create generators from the ``source code" of a specification. Unfortunately, this problem is NP-Hard in even simple cases. If we limit ourselves to preconditions $\phi$ which consist only of arithmetic constraints (which we will in fact do), the problem of generating hundreds of sets of inputs which satisfy $\phi$ is as hard as generating \textit{one} set of inputs, which is equivalent to SMT solving.

@@

Then, to generate a concretization ordering, we randomly depth-first search $G(P)$, and list variables in the order that they're visited in the graph. Pseudocode for this is shown in \autoref{alg:generate-ordering}.

\begin{algorithm}
    \caption{Generate a Random Concretization Ordering}
    \label{alg:generate-ordering}
    \begin{algorithmic}
      \Function{randomConcr}{$P$}
      \State $G \gets G(P)$ 
      \State $X \gets \{\}$
      \State $S \gets $ shuffle(V($G$))
      \State $xs \gets []$

      \While{$S \neq []$}
       \State $x \gets $ Pop($S$)
       \If{$x \notin X$}
         \State $X \gets X \cup \{x\}$
         \State $xs \gets $ append($xs$,$[x]$)
         \State $S \gets $ append(shuffle(N($x$)),$S$)
       \EndIf

      \EndWhile

      \Return $xs$

      \EndFunction
    \end{algorithmic}
\end{algorithm}

Then, to generate our set of generator candidates, we repeatedly run this function. This may give repeated generators, and so we filter the result for uniqueness. It's worth noting that getting repeated results is good: it means that we have successfully pruned the search space sufficiently to get a small number of possible generators. The number of generators in our set requires a careful balance. Too few and the set may not contain a generator which succeeds often enough to rapidly generate our desired number of unique inputs. Too many and the learning algorithm will converge too slowly to the best generator in the set. Empirically, we have found that $n^2$ (where $n$ is the number of variables in $P$) is a good number of generators to take.

@@

In \autoref{tab:data}, we show the results for some trials of our algorithm. In each trial, we run the generration algorithm on the specified constraint five times for $T = 1000$ iterations. $K$ is the average number of unique scripts discovered by \autoref{alg:generate-ordering}, Rejected is the average number of failed generator samples that happen during the runs, and Unique is the average number of unique input sets generated. In all cases, the generation takes less than 1ms of wall clock time, as reported by the Haskell library \texttt{timeit}. By default, integers are generated in the range $[-10000,10000]$ instead of the full integer range.

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Making the Best of a Bad Situation}
\subtitle{Picking Random Generators with Multi-Armed Bandits}

@@

\section{Motivation}
% Deductive Program Verifiers like Dafny and Frama-C have proven themselves both
% useful and powerful tools in the fight against un-verified software.  However,
% their adoption is hindered by the fact that graduate-level expertise in
% verification is required in order to verify any non-trivial program. One
% approach to lowering the barrier to entry for deductive verifiers is to
% incorporate lighter-weight formal methods into the mix. While SMT based
% verification is very powerful, its nuances lead to much of the difficulty in
% using verifiers based on it, and so selectively substituting this verification
% technique with a less precise but easier-to-use formal method would lower the
% difficulty while retaining the same workflow and some of the theoretical
% guarantees.

In this extended abstract, we propose and investigate the use of \textit{Property-Based Testing} in
place of SMT solving as a core for program verification tools like Dafny and Frama-C.
Property-Based Testing (PBT) \cite{qc} is a lightweight formal method which checks the
specifications for a program by randomly generating thousands of
inputs, and ensuring that the program meets its specification at each concrete value.

A PBT-based approach to program verification has significant usability benefits
over SMT-based verification. While using SMT-based verifiers, it is often difficult
to interpret the SMT output. When a solver fails to verify a proposition,
either (a) the specification is incorrect, or
(b) the solver was simply unable to find a proof. These two results are often
indistinguishable, and so the only course of action is
to attempt to provide more hints to the solver. With PBT, the only failure mode is through
counter-examples, and so no interpretation is required \footnote{
The counter-examples in a PBT approach are also more easily interpretable. With SMT,
counter-models are encoded in an SMT data format and are not easily lifted
back to program values.
}. PBT can also aid in
the development of SMT-verified programs by allowing the user to test
specifications to gain confidence that they are true before spending costly time
attempting to develop tricky loop invariants.

Moreover, PBT is fully modular, and so it can be used selectively:
randomly-tested code and SMT-verified code can coexist in the same file,
enabling users to take a \textit{gradual verification} approach by gradually
replacing tested code with verified code.

In this paper, we develop a simple deductive program verifier
which is backed by PBT instead of SMT. In \autoref{sec:pbt}, we show how PBT can
be used in place of an SMT solver to do deductive verification. This process is
conceptually simple, but we will see that it hides a major challenge: generating
thousands of independent random input sets which satisfy a function's
precondition is NP-hard. In fact, tackling a small variant of problem is the
bulk of the content of this report, and its main technical contribution. We give
a sketch of this contribution in this section. In brief, our technique first
infers a set of ``candidate" example-generators from the precondition in
question, and then uses a reinforcement-learning-based online learning algorithm
to find the best generator among the set.

@@

@@

While 
In terms of ($\ast$), this means that a user of a verifier backed by PBT would
have to provide programs which generate random heaps $\sigma$ to satisfy the
preconditions of every function they plan to test. This is labor-intensive,
error prone, and diminishes the benefits of PBT over SMT-based verification. To
make the former practical, we need to \textit{automatically} create generators
from the ``source code" of a specification. Unfortunately, this problem is
NP-Hard in even simple cases. If we limit ourselves to preconditions $\phi$
which consist only of arithmetic constraints (which we will in fact do), the
problem of generating hundreds of sets of inputs which satisfy $\phi$ is as hard
as generating \textit{one} set of inputs, which is equivalent to SMT solving.

@@

Then, to generate a concretization ordering, we randomly depth-first search $G(P)$, and list variables in the order that they're visited in the graph.
% Pseudocode for this is shown in \autoref{alg:generate-ordering}.

% \begin{algorithm}
%     \caption{Generate a Random Concretization Ordering}
%     \label{alg:generate-ordering}
%     \begin{algorithmic}
%       \Function{randomConcr}{$P$}
%       \State $G \gets G(P)$ 
%       \State $X \gets \{\}$
%       \State $S \gets $ shuffle(V($G$))
%       \State $xs \gets []$

%       \While{$S \neq []$}
%        \State $x \gets $ Pop($S$)
%        \If{$x \notin X$}
%          \State $X \gets X \cup \{x\}$
%          \State $xs \gets $ append($xs$,$[x]$)
%          \State $S \gets $ append(shuffle(N($x$)),$S$)
%        \EndIf

%       \EndWhile

%       \Return $xs$

%       \EndFunction
%     \end{algorithmic}
% \end{algorithm}

To generate our set of generator candidates, we repeatedly run this function. This may give repeated generators, and so we filter the result for uniqueness. It's worth noting that getting repeated results is good: it means that we have successfully pruned the search space sufficiently to get a small number of possible generators. The number of generators in our set requires a careful balance. Too few and the set may not contain a generator which succeeds often enough to rapidly generate our desired number of unique inputs. Too many and the learning algorithm will converge too slowly to the best generator in the set. Empirically, we have found that $n^2$ (where $n$ is the number of variables in $P$) is a good number of generators to take.

@@

In \autoref{tab:data}, we show the results for some trials of our algorithm. In each trial, we run the generration algorithm on the specified constraint five times for $T = 1000$ iterations. $K$ is the average number of unique scripts discovered by the random DFS, Rejected is the average number of failed generator samples that happen during the runs, and Unique is the average number of unique input sets generated. In all cases, the generation takes less than 1ms of wall clock time, as reported by the Haskell library \texttt{timeit}. By default, integers are generated in the range $[-10000,10000]$ instead of the full integer range.
